---
title: "Paper X - Causal Roadmap"
keywords:
  - Epidemiology
  - Causal Inference
  - Triangulation
metadata-files:
  - ../../../all_phd/authors.yml
format:
  pdf:
    pdf-engine: pdflatex
    number-sections: true
    toc: false
    margin-left: 15mm
    margin-right: 15mm
    margin-bottom: 20mm
    margin-top: 15mm
    include-in-header:
      - \usepackage{rotating}
      - \usepackage{float}
      - \usepackage{typearea}
    pandoc_args:
      - !expr acronymsdown::add_filter()
bibliography: ../../../PhD.bib
link-citations: true
linkcolor: blue
execute:
  echo: false
  warning: false
  error: false
  message: false
editor_options: 
  chunk_output_type: console
filters:
  - /home/lorenzo/R/x86_64-pc-linux-gnu-library/4.1/acronymsdown/parse-acronyms.lua
  - ../../../all_phd/scholarly-metadata.lua
  - ../../../all_phd/author-info-blocks.lua
acronyms:
  fromfile: ../../../all_phd/acronyms.yml
  insert_links: false
  insert_loa: false
  sorting: alphabetical
---

```{=latex}
\setcounter{tocdepth}{4}
\tableofcontents
```

```{r}
#| include: false

here::i_am("causal_roadmap.qmd")
source("../../../all_phd/info_phd.R")
source("../../../all_phd/triangulation_methods.R")
infos <- info_phd()
methods <- describe_triangulation_methods()

library(tidyverse)
```

```{r}
# Type of adjustment set from DAG
type_mas <- "minimal"
# Type of effect to be estimated from DAG
type_effect <- "direct"

# Number of folds for Cross-Validation (CV)
cv.folds <- 10
```

\newpage

Most of the following was **copy-pasted** from published papers.

<!-- ####################################################################### -->
# Formulate the research question(s)
<!-- ####################################################################### -->

<!-- ####################################################################### -->
## Descriptive analyses
<!-- ####################################################################### -->

{{< include ../../../all_phd/descriptive_stats.qmd >}}

<!-- ####################################################################### -->
# Define a realistic statistical model
<!-- ####################################################################### -->

<!-- ####################################################################### -->
# Specify a causal model and causal quantity of interest
<!-- ####################################################################### -->

```{r}
rq1.title <- ""
```

## Research question 1: `r rq1.title`

<!-- ####################################################################### -->
# Identification and the statistical estimand
<!-- ####################################################################### -->

<!-- ####################################################################### -->
# Estimation from data and statistical inference
<!-- ####################################################################### -->

<!-- ####################################################################### -->
# Triangulation: Improving causal inference in aetiological epidemiology
<!-- ####################################################################### -->

{{< include ../../../all_phd/tab_triangulation_pX.qmd >}}

<!-- ####################################################################### -->
# Interpretation and sensitivity analyses to inform a substantive conclusion
<!-- ####################################################################### -->

## Reliable Causal Inference

{{< include ../../../all_phd/app-rel-ci.qmd >}}

## Reliable Statistical Inference

{{< include ../../../all_phd/app-rel-si.qmd >}}

## Replicability

{{< include ../../../all_phd/app-replicability.qmd >}}

<!-- ####################################################################### -->
# Appendix
<!-- ####################################################################### -->

## Data Checking

{{< include ../../../all_phd/app-data-checking.qmd >}}

## Checklist A: Replicability

Table with random seed, names, description, and version numbers of all software packages.

| **Name** | **Version** | **Description** |
|:---------|:------------|:----------------|
| Random seed | NA | Will be set to $X$ |
| `R` | `r I(paste0(R.version$major, ".", R.version$minor))` | Statistical programming environment |
| ... | ... | ... |

## Checklist B1: `xxx` package specifications

Table providing values for all non-data arguments, and brief rationale when departing from the default specification.

| **Argument** | **Setting** | **Default** (Y/N) | **Comment** |
|:-------------|:------------|:------------------|:------------|
| ... | ... | ... | ... |

## DAGs

```{r}
source("../DAGs/dag_v1.R")
source("../code/dags.R")

dags <- dag_v1()
```

### Research question 1: `r rq1.title`

<!-- ####################################################################### -->
# References {.unnumbered}
<!-- ####################################################################### -->

::: {#refs}
:::
